
# Data Processing and Mapping

```{r}
source("setup.R")
```

## Read in rasters

From L: Drive
```{r}
# read in raw files from L drive:
hfi_1999 <- terra::rast("L:\\Projects_active\\2024_x2025_Human_Footprint_Index\\data\\1999_int16.tif")


hfi_2023 <- terra::rast("L:\\Projects_active\\2024_x2025_Human_Footprint_Index\\data\\2023_int16.tif")

hfi_change <- terra::rast("L:\\Projects_active\\2024_x2025_Human_Footprint_Index\\data\\2023m1999_int16.tif")

```

Local data/ folder (on Mac)
```{r}
hfi_1999 <- terra::rast("data/1999_int16.tif")
```


## Projection

Project for Leaflet
```{r}
# this will take a while...1 hour on Mac
system.time(
hfi_1999_projected <- terra::project(hfi_1999, "EPSG:3857")
)

writeRaster(hfi_1999_projected, "data/hfi_1999_leaflet.tif")

```


## Aggregate to ~ 1km, project and save
```{r}
# aggregate
hfi_change_1km <- terra::aggregate(hfi_change, fact = 3, fun = mean)

# project; the output resolution is 464km....
hfi_change_1km_leaflet <- terra::project(hfi_change_1km, "EPSG:3857")


# save
writeRaster(hfi_change_1km_leaflet, "L:\\Projects_active\\2024_x2025_Human_Footprint_Index\\data/2023_1999_900m_leaflet.tif")
```



## Tile Creation

Had to run this in terminal after installing gdal (Mac)
```{bash}
gdal_translate -of VRT -ot Byte -scale /Users/caitlinmothes/Desktop/HFI_shiny/data/1999_int16.tif temp.vrt

gdal2tiles temp.vrt
```

Files default saved in Users/caitlinmothes, moved to data/ folder. Folder is 3GB in size though for zoom level 9...


## Leaflet Mapping

Tiles Josh hosted on AGOL
```{r}
# ArcPro Map Tiles ---------
# 
# url <- "https://tiles.arcgis.com/tiles/swlKRWoduvVuwMcX/arcgis/rest/services/TP_1999_3857/MapServer/tile/{z}/{y}/{x}"

# new URL
url <- "https://tiles.arcgis.com/tiles/swlKRWoduvVuwMcX/arcgis/rest/services/TP_1999_3857_12levels/MapServer/tile/{z}/{y}/{x}"

url2 <- "https://tiles.arcgis.com/tiles/swlKRWoduvVuwMcX/arcgis/rest/services/TP_Change_2023_1999_1k/MapServer/tile/{z}/{y}/{x}"


leaflet() %>% 
  addTiles() %>% 
  addTiles(url,
           options = tileOptions(maxNativeZoom = 12))
```


# Read in Vector data

```{r}
# Load global country borders

world <- ne_countries(returnclass = "sv", scale = "large")


# save to data/ folder
writeVector(world, "data/countries.shp", overwrite = TRUE)


# Import IPCC A5 Reference Regions

# Download the ZIP file
download.file("https://www.ipcc-data.org/documents/ar5/regions/referenceRegions.zip", "data/IPCC_regions.zip", mode = "wb") 

# Unzip the file
unzip("data/IPCC_regions.zip", exdir = "data/IPCC_regions")

# remove the .zip file
file.remove("data/IPCC_regions.zip")



```

Read the files in
```{r}
countries <- terra::vect("data/countries.shp")

ipcc <- terra::vect("data/IPCC_regions/referenceRegions.shp")
```


Add to map
```{r}

countries_sf <- st_as_sf(countries)
  
leaflet() %>% 
  addTiles() %>% 
  addPolygons(data = countries_sf, color = "blue", fillColor = "red", weight = 1, fillOpacity = 0, popup = ~name, group = "Countries") %>% 
  addLayersControl(
    overlayGroups = c("Countries", "Borders"),  # Specify the layers to toggle
    options = layersControlOptions(collapsed = FALSE)  # Control options
  )
```


# Calculate zonal stats

```{r}

# Create a blank raster matching the resolution/extent of r
poly_raster <- rast(hfi_1999)


# Rasterize the polygon vector based on IDs
ipcc_poly <- rasterize(ipcc, poly_raster, field = "NAME")

country_poly <- rasterize(countries, poly_raster, field = "name")

# 1.5 minutes
system.time(
  zonal_ipcc <- terra::zonal(hfi_1999, ipcc_poly, fun = "mean", na.rm = TRUE)
)

system.time(
  zonal_country <- terra::zonal(hfi_1999, country_poly, fun = "mean", na.rm = TRUE)
)
```

Calculate over all years
```{r}

# Create a blank raster matching the resolution/extent of r
poly_raster <- rast(hfi_1999)


# Rasterize the polygon vector based on IDs
ipcc_poly <- rasterize(ipcc, poly_raster, field = "NAME")

country_poly <- rasterize(countries, poly_raster, field = "name")


zonal_ipcc_all <- map(list(hfi_1999, hfi_2023), ~ terra::zonal(.x, ipcc_poly, fun = "mean", na.rm = TRUE))

zonal_country_all <- map(list(hfi_1999, hfi_2023), ~ terra::zonal(.x, country_poly, fun = "mean", na.rm = TRUE))

```

Add to polygons and save
```{r}
# countries
country_all <- reduce(zonal_country_all, left_join, by = "name")

countries <- merge(countries, country_all, by = "name")

writeVector(countries, "app/app_data/countries.shp", overwrite = TRUE)

# ipcc regions
ipcc_all <- reduce(zonal_ipcc_all, left_join, by = "NAME")

ipcc <- merge(ipcc, ipcc_all, by = "NAME")

writeVector(ipcc, "app/app_data/IPCC_regions/referenceRegions.shp", overwrite = TRUE)
```


Remove water pixels
```{r}
mask_1999 <- terra::mask(hfi_1999, countries)
```

